---
title: "Assignment 7"
author: "Group 26"
format: html
editor: visual
---

## Load libraries

```{r}
library(tidyverse)
library(broom)
```

## Load data

```{r}

#| eval: false

raw_dir <- "data/_raw/"
data_file <- "gravier.RData"
data_loc <- "https://github.com/ramhiser/datamicroarray/raw/master/data/"

if( !dir.exists("data") ){
  dir.create(path = "data")
}
if( !dir.exists(raw_dir) ){
  dir.create(path = raw_dir)
}
if( !file.exists(str_c(raw_dir, data_file)) ){
  download.file(
    url = str_c(data_loc, data_file),
    destfile = str_c(raw_dir, data_file))
}
load(file = str_c(raw_dir, data_file))

```

We want to perform PCA, so we want to do these three things to make an analysis:

1) Look at the data in PC coordinates.
2) Look at the rotation matrix.
3) Look at the variance explained by each PC.

### Look at the data in PC coordinates

```{r}
# Make dataset tidy
gravier <- gravier |>
  bind_cols() |>
  as_tibble() |> 
  relocate(Early_metastasis = y)

# Fit the PCA
pca_fit <- gravier |>  
  drop_na() |> 
  select(where(is.numeric)) |>  # retain only numeric columns
  prcomp(scale = TRUE) # do PCA on scaled data

#Plot PCA
pca_fit |> 
  augment(gravier) |>  # add original dataset back in
  ggplot(aes(x = .fittedPC1, 
             y  = .fittedPC2, 
             color = Early_metastasis)) + 
  geom_point() +
  scale_color_manual(
    values = c(good = "#D55E00", 
               poor = "#0072B2")) +
  labs( title = "PCA plot of Gravier data",
           x = "PCA1",
           y = "PCA2") +
  theme(plot.title = element_text(hjust = 0.5)) +
  theme_minimal()
```
This is a visualization of the data in the first two principal component coordinates. These components are linear combinations of the original variables. The plot helps explore how the data points are distributed in a reduced-dimensional space defined by the first two principal components.

The Early_metastasis helps to see if there is any separation or clustering of data points based on early metastasis status. In the previous plot we can't see aparently any separation based on early metastasis, showing that the groups "poor" and "good" have different patterns in the original data.


```{r}
gravier |>  View()
pca_fit |> View()
str(pca_fit)
```

### Plot the rotation matrix

```{r}

# define arrow style for plotting
arrow_style <- arrow(
  angle = 20, ends = "first", type = "closed", length = grid::unit(8, "pt")
)

# Extract information in rotation matrix and plot it 
pca_fit |> 
  tidy(matrix = "rotation") |> 
  pivot_wider(names_from = "PC", names_prefix = "PC", values_from = "value") |> 
  ggplot(aes(PC1, PC2)) +
  geom_segment(xend = 0, yend = 0, arrow = arrow_style) +
  geom_text(
    aes(label = column),
    hjust = 1, vjust=1 , nudge_x = 0, 
    color = "pink"
  ) +
  xlim(-0.075, .075) + ylim(-.05, 0.05) +
  coord_fixed()+  # fix aspect ratio to 1:1
  theme_minimal()
 

```

### Variance explained by each PC

```{r}

pca_fit |> 
  tidy(matrix = "eigenvalues")
#Extract the varience information and plot the percentage variance explained by each PC
pca_fit |> 
  tidy(matrix = "eigenvalues") |> 
  ggplot(aes(PC, percent)) +
  geom_col(fill = "#56B4E9", 
           alpha = 0.8) +
  scale_x_continuous(breaks= waiver()) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.01))
  ) + 
  theme_minimal()
```

So what exactly does all this mean?
